---
title: "Income Support Caseload Forecast: Technical Appendix"
subtitle: "Supporting Evidence for Note-1 Briefing — March 2026"
author: "Strategic Data Analytics Team"
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    toc-depth: 3
    code-fold: show
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
    keep-md: false
editor: visual
editor_options:
  chunk_output_type: console
---

> **How to use this document.** This appendix substantiates every claim made in the companion briefing note (`note-1-briefing.pdf`). Sections are organized to mirror the briefing's structure. Use the table of contents to navigate directly to the evidence behind a specific briefing claim.

------------------------------------------------------------------------

```{r}
#| label: reconcile-workdir
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir = '../../')
```

```{r}
#| label: set-options
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
report_render_start_time <- Sys.time()
opts_chunk$set(
  results      = 'show',
  message      = FALSE,
  warning      = FALSE,
  comment      = NA,
  tidy         = FALSE,
  out.width    = "960px",
  fig.width    = 10,
  fig.height   = 6,
  fig.path     = 'figure-png-iso/'
)
echo_chunks    <- FALSE
message_chunks <- FALSE
options(width = 100)
read_chunk("analysis/note-1/note-1-appendix.R")
```

```{r}
#| label: load-packages
#| code-summary: Packages used in this report
#| echo: false
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-sources
#| echo: false
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-globals
#| echo: false
#| results: hide
#| message: false
#| code-fold: true
#| code-summary: Paths, config parameters, colour constants
```

```{r}
#| label: declare-functions
#| echo: false
#| results: hide
#| message: false
#| code-fold: true
#| code-summary: Helper functions (fiscal year label, fmt_num, fmt_pct)
```

```{r}
#| label: load-data
#| echo: false
#| results: hide
#| message: false
#| code-fold: true
#| code-summary: Load forecast artifacts from Lanes 3–5
```

```{r}
#| label: tweak-data-1
#| echo: false
#| results: hide
#| message: false
#| code-fold: true
#| code-summary: Parse dates, create tier_label_display, residuals
```

------------------------------------------------------------------------

# 1 Data Foundation

*Supports briefing section "Interpretation Guidance — The forecast is not policy-sensitive."*

## 1.1 Data Source and Coverage

The forecast pipeline is built on the publicly available [Alberta Income Support Aggregated Caseload Data](https://open.alberta.ca/opendata/income-support-aggregated-caseload-data), downloaded from Alberta Open Data via **Lane 1 (Ferry)**. The analysis-ready time series spans **246 consecutive months** from April 2005 to September 2025, the largest dimension available across all data tables.

```{r}
#| label: provenance-dashboard
#| echo: false
#| code-fold: true
#| code-summary: Manifest summary (forge_hash, train/test split, lineage check)
```

## 1.2 Dimension Availability Phases

Different demographic breakdowns were added to the program's reporting at different times. The table below shows which dimensions are available for the forecast. Only **Total Caseload** spans the full 20-year history and is used by the Tier 1 and Tier 2 models. Client-type disaggregation (Tier 3, planned) requires April 2012 onward.

| Phase |       Period        | Months | Dimensions Available               |
|:-----:|:-------------------:|:------:|:-----------------------------------|
|   1   | Apr 2005 – Mar 2012 |   84   | Total caseload only                |
|   2   | Apr 2012 – Mar 2018 |   72   | \+ Client type, Family composition |
|   3   | Apr 2018 – Mar 2020 |   24   | \+ ALSS Regions                    |
|   4   | Apr 2020 – Mar 2022 |   24   | \+ Age groups, Gender (binary)     |
|   5   | Apr 2022 – Sep 2025 |   42   | \+ Gender (non-binary)             |

## 1.3 Historical Context: Structural Breaks

The most important context for interpreting forecast uncertainty is the history of structural breaks — external shocks that produced large, rapid deviations from any statistical model's prediction.

```{r}
#| label: g4
#| fig.height: 6
#| code-fold: true
#| code-summary: Figure — 20-year history with annotated structural breaks
```

Four major structural breaks are annotated above:

-   **2008 Financial Crisis** — Sharp caseload increase as unemployment rose sharply across Alberta.
-   **2012 Alberta Works Reform** — Policy change restructured benefit delivery; administrative reclassification created apparent step-change in reported counts.
-   **2014 Oil Price Collapse** — Sustained caseload growth lasting approximately three years as resource-sector employment contracted.
-   **2020 COVID-19 Pandemic** — Largest and fastest shock in the 20-year record: caseload peaked above 80,000 before rapidly normalising as federal support programs were introduced and wound down.

Each structural break represents a scenario where actual caseloads deviated substantially from any reasonable statistical extrapolation. These events are why prediction intervals widen so dramatically at 24 months.

## 1.4 Data Quality Notes

-   **Nov 2020 duplication**: The source data contained a duplicate row for November 2020 family composition; the corrected row (44,850 total) was retained in Ellis lane processing.
-   **Small-count suppressions**: Six cells across client-type tables were suppressed in the source (counts below reporting threshold); retained as `NA` in analysis tables.
-   **Regions renaming**: The "North Central" region was renamed to "Unknown" effective January 2022; reason not documented in source.

------------------------------------------------------------------------

# 2 The Forecast

*Supports briefing section "Key Findings."*

## 2.1 Hero Forecast Chart

The chart below presents the full 20-year historical series together with the 24-month ARIMA projection. Colors follow GoA corporate palette (sky-mid for history, goateal for forecast, sky-light for intervals).

```{r}
#| label: g1
#| fig.height: 6
#| code-fold: true
#| code-summary: Figure — 20-year history + 24-month ARIMA forecast with prediction intervals
```

## 2.2 Full 24-Month Forecast Table

The table below shows every month of the ARIMA forecast along with 95% prediction intervals and the absolute and percentage change relative to the September 2025 baseline of **`r scales::comma(current_caseload)`** cases.

```{r}
#| label: t1
#| code-fold: true
#| code-summary: "> Key forecast numbers: ARIMA at +6, +12, +18, +24 months"

t1_data %>%
  kable(
    caption = paste0(
      "ARIMA (3,1,1)(1,0,0)[12] forecast at key horizons. ",
      "Baseline: ", scales::comma(current_caseload), " cases (Sep 2025). ",
      "All values on original (count) scale, back-transformed from log."
    ),
    align = c("r", "l", "l", "r", "l", "r", "r", "r")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  column_spec(4, bold = TRUE) %>%
  column_spec(5, color = "gray50") %>%
  footnote(general = paste0(
    "PI Width = 95% PI High minus 95% PI Low. ",
    "At +6 months PI Width ≈ 16,000 cases; at +24 months PI Width ≈ 61,000 cases."
  ))
```

**Reading the table**: the central forecast is essentially flat (all "Change (%)" values under ±2%). Leadership should not interpret stability as certainty — it means the model's best statistical guess, given the current trend, is that caseload will not grow materially. External shocks would override this.

------------------------------------------------------------------------

# 3 Model Performance

*Supports briefing section "Model Confidence."*

## 3.1 Model Specifications

| Field                | Tier 1: Seasonal Naive            | Tier 2: ARIMA                                               |
|:-----------------------|:-----------------------|:-----------------------|
| **Type**             | Rule-based                        | Statistical                                                 |
| **Specification**    | Repeat last 12 months             | (3,1,1)(1,0,0)\[12\]                                        |
| **Data transform**   | None                              | Log (variance stabilisation)                                |
| **Training data**    | Full series (Apr 2005 – Sep 2023) | Full series (Apr 2005 – Sep 2023)                           |
| **Selection method** | Fixed rule                        | `forecast::auto.arima(stepwise=FALSE, approximation=FALSE)` |
| **AIC / BIC**        | N/A                               | −1418.4 / −1397.4                                           |
| **Parameters**       | 0                                 | 6                                                           |

The ARIMA order (3,1,1)(1,0,0)\[12\] means:

-   **(p=3, d=1, q=1)** — Autoregressive lag-3, first-differenced to remove trend, moving average lag-1
-   **(P=1, D=0, Q=0)\[12\]** — Seasonal autoregressive lag-12 (annual cycle), no seasonal differencing, no seasonal MA
-   **Log transform** applied before model fitting; back-transformed (exp) after forecast generation

The differencing (d=1) confirms what Augmented Dickey-Fuller tests showed in EDA-2: the caseload series is non-stationary in levels but stationary after first-differencing. The log transform stabilises the variance, preventing the model from producing wider intervals during periods of higher caseload.

## 3.2 Backtest Performance Metrics

```{r}
#| label: t2
#| code-fold: true
#| code-summary: "> Model performance table (RMSE, MAE, MAPE)"

t2_data %>%
  kable(
    caption = paste0(
      "Backtest performance: 24-month hold-out window (Oct 2023 – Sep 2025). ",
      "All metrics on original count scale (back-transformed from log). ",
      "n_test = 24, n_train = 222."
    ),
    align = c("l", "l", "r", "r", "r")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(2, bold = TRUE, background = "#f0f8ff") %>%
  footnote(
    general = paste0(
      "RMSE = root mean squared error (penalises large errors). ",
      "MAE = mean absolute error. ",
      "MAPE = mean absolute percentage error (scale-free). ",
      "Lower values indicate better backtest accuracy. ",
      "ARIMA reduces MAPE by 3.1 percentage points vs Naive, ",
      "equivalent to ~1,700 fewer cases per month average error."
    )
  )
```

## 3.3 Model Comparison: Divergence Over the Forecast Horizon

```{r}
#| label: g21
#| fig.height: 6
#| code-fold: true
#| code-summary: Figure — ARIMA vs Naive overlaid (divergence chart)
```

The ARIMA model (teal) and the Seasonal Naive model (orange) start close together at the beginning of the forecast window and diverge over time. The Naive model anchors to the October 2024 – September 2025 seasonal pattern, which recently trended slightly lower (\~57,000-60,000). The ARIMA model captures the momentum of the recent stabilisation at slightly higher levels (\~61,000). The gap between them widens from approximately 3,600 cases at month 1 to 4,200 cases at month 24.

## 3.4 Faceted Forecast: Side-by-Side Comparison

```{r}
#| label: g2
#| fig.height: 8
#| code-fold: true
#| code-summary: Figure — Faceted forecast (one panel per model tier, shared y-axis)
```

The faceted view uses a shared y-axis to preserve comparability of absolute forecast levels. Key observations:

-   The Naive model's prediction intervals are symmetric and constant-width (reflecting its rule-based variance estimate).
-   The ARIMA model produces slightly narrower intervals at short horizons (benefit of log-transform variance stabilisation) but wider intervals at long horizons (reflecting propagated AR uncertainty).

------------------------------------------------------------------------

# 4 Backtest Evidence

*Supports briefing section "Model Confidence — 13.3% MAPE."*

## 4.1 Actual vs Model-Fitted

The charts below compare the model-fitted values against the actual observed caseload over the 24-month backtest window (October 2023 – September 2025). This period was held out during training — the models never saw it.

```{r}
#| label: g3
#| fig.height: 8
#| code-fold: true
#| code-summary: Figure — Actual vs fitted for both model tiers
```

Observations:

-   **Tier 2 ARIMA** tracks month-to-month movements more closely, reflecting learned autoregressive structure (the model uses recent history to anticipate direction).
-   **Tier 1 Naive** follows the general seasonal shape but is anchored to the prior year's level. During the early backtest period this over-estimated because 2022-23 was unusually high post-COVID.

## 4.2 Residual Analysis

```{r}
#| label: g31
#| fig.height: 8
#| code-fold: true
#| code-summary: Figure — Residuals (actual minus fitted) by month
```

The residual pattern reveals model bias:

-   **Tier 1 Naive**: Predominantly negative residuals in the second half of the backtest window, indicating systematic over-prediction. The Naive model carried the higher post-COVID caseload forward while actuals normalised downward.\
-   **Tier 2 ARIMA**: Residuals are more balanced around zero with no clear systematic bias, indicating the ARIMA model adapted more quickly to the post-COVID normalisation.

## 4.3 Percentage Error by Month

```{r}
#| label: g32
#| fig.height: 8
#| code-fold: true
#| code-summary: Figure — Absolute percentage error vs MAPE reference line
```

The dashed line marks each model's MAPE. Months above the line performed worse than average for that tier. Both models show higher errors at fiscal year transitions (April–May), which likely reflects administrative processing effects in the source data not captured by statistical patterns alone.

------------------------------------------------------------------------

# 5 Methodology Notes

*Supports briefing section "Model Confidence — how the model was built."*

## 5.1 Train / Test Split Strategy

The pipeline uses a **24-month hold-out** design:

-   **Training window**: April 2005 – September 2023 (222 months)
-   **Test / backtest window**: October 2023 – September 2025 (24 months, never used during training)
-   **Forecast window**: October 2025 – September 2027 (24 months into the future)

The backtest window is equal in length to the forecast horizon. This gives an honest estimate of how well the model would have performed had it been deployed 24 months ago — the most direct analogue to the real-world forecasting task.

## 5.2 Log Transformation

The series is log-transformed before model fitting and back-transformed (exponentiated) for reporting. The decision was codified in the Mint lane based on EDA-2 findings:

1.  **Variance stabilisation**: caseload variance scales with level; log-transforming makes variance approximately constant across the series, a key ARIMA assumption.
2.  **Multiplicative seasonality**: the log transform converts multiplicative seasonal patterns into additive ones, making them easier for ARIMA to capture.
3.  **Forecast intervals**: back-transforming produces asymmetric intervals on the original scale (PI high − point \> point − PI low), which is appropriate for a count variable that cannot go below zero.

## 5.3 Model Selection Criterion

`forecast::auto.arima(stepwise = FALSE, approximation = FALSE)` was used to select the ARIMA order. This performs an exhaustive search over candidate models, selecting the specification with the lowest **AICc** (Akaike Information Criterion corrected for finite samples). The selected order ARIMA(3,1,1)(1,0,0)\[12\] balances model fit against parsimony.

## 5.4 Pipeline Lineage

Every forecast artifact is linked by a `forge_hash` — a SHA-256 digest of the Mint lane's data contract — ensuring that the forecast and the training data can never become silently misaligned.

```{r}
#| label: lineage-table
#| echo: false
#| code-fold: true
#| code-summary: Manifest chain summary

lineage_df <- data.frame(
  Lane = c("Lane 3 (Mint)", "Lane 4 (Train)", "Lane 5 (Forecast)"),
  Artifact = c("forge_manifest.yml", "model_registry.csv", "forecast_manifest.yml"),
  Hash_Field = c("forge_hash", "forge_hash (consumed)", "forge_hash_consumed + forecast_hash"),
  Value = c(
    substr(forge_manifest$forge_hash, 1, 12),
    substr(model_registry$forge_hash[1], 1, 12),
    substr(forecast_manifest$forge_hash_consumed, 1, 12)
  )
)

lineage_df %>%
  kable(caption = "Pipeline lineage chain (first 12 characters of each hash shown).") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
  footnote(general = paste0(
    "When forge_hash in forecast_manifest equals forge_hash in forge_manifest, ",
    "the forecast is confirmed to derive from the current training data. ",
    "Hash integrity check: ",
    ifelse(
      forecast_manifest$forge_hash_consumed == forge_manifest$forge_hash,
      "[PASS] Lineage intact.",
      "[FAIL] Mismatch detected — re-run pipeline."
    )
  ))
```

------------------------------------------------------------------------

```{r}
#| label: render-time
#| echo: false
elapsed <- round(as.numeric(difftime(Sys.time(), report_render_start_time, units = "secs")), 1)
```

*Appendix rendered in `r elapsed` seconds. R version `r R.version$major`.`r R.version$minor`. Focal date: `r focal_date`. forecast_hash: `r substr(forecast_manifest$forecast_hash, 1, 16)`...*

*AI Assistance Disclosure: This document was generated with the assistance of GitHub Copilot (Claude). All analytical content reflects the author's professional judgment and was reviewed against pipeline output artifacts.*