---
title: "Alberta Income Support: Caseload Forecast Report"
subtitle: "24-Month Forward Projections — October 2025 to September 2027"
author: "Strategic Data Analytics Team"
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: show
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
    keep-md: false
editor: visual
editor_options: 
  chunk_output_type: console
---

# Executive Summary

Alberta's Income Support caseload is projected on a modest upward trajectory through September 2027 under the preferred ARIMA model. Uncertainty bands are wide — the 95% prediction interval spans several thousand cases at the 24-month horizon — reflecting genuine model uncertainty in the face of historical structural breaks (2008 recession, 2014 oil price shock, COVID-19). Exact point-forecast numbers are presented in the **Key Forecast Numbers** table.

Two forecasting models are available in this phase: a **Seasonal Naive baseline** (Tier 1) and an **ARIMA model** (Tier 2). The ARIMA model substantially outperforms the naive baseline on every accuracy metric over the 24-month held-out backtest window. Future phases will add Tier 3 (subgroup disaggregation: per-client-type forecasting, then summed) and Tier 4 (ARIMA with time-varying economic predictors), which are expected to reduce forecast uncertainty.

The analysis pipeline is fully reproducible and versioned. All outputs trace back to a unique `forge_hash` linking the forecast to the exact historical data slice used for model training.

---

```{r}
#| code-summary: Reconciles working directory between execution modes
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir='../../')
```

```{r}
#| label: set_options
#| code-summary: Report-wide chunk and display options
#| echo: false
#| results: hide
#| message: false
cat("Working directory: ", getwd())
report_render_start_time <- Sys.time()
opts_chunk$set(
  results      = 'show',
  message      = FALSE,
  warning      = FALSE,
  comment      = NA,
  tidy         = FALSE,
  out.width    = "960px",
  fig.width    = 10,
  fig.height   = 6,
  fig.path     = 'figure-png-iso/'
)
echo_chunks    <- FALSE
message_chunks <- FALSE
options(width = 100)
ggplot2::theme_set(ggplot2::theme_minimal())
read_chunk("analysis/report-1/report-1.R")
```

## The Forecast

The chart below presents the full 20-year historical series together with the 24-month ARIMA projection. The vertical dashed line marks **September 2025**, the latest month with observed data. Everything to the right is a model-generated prediction.

```{r}
#| label: load-packages
#| code-summary: Packages used in this report
#| echo: false
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-sources
#| code-summary: Custom functions from project scripts
#| echo: false
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-globals
#| code-summary: Paths, config, colour constants
#| echo: false
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-functions
#| code-summary: Local helper functions
#| echo: false
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-data
#| code-summary: Load forecast artifacts from Lanes 3–5
#| echo: false
#| results: hide
#| message: false
#| code-fold: true
#| cache: true
```

```{r}
#| label: tweak-data-1
#| code-summary: Parse dates, factor labels, residual columns
#| echo: false
#| results: hide
#| message: false
#| code-fold: true
#| cache: true
```

```{r}
#| label: g1-data-prep
#| code-summary: Bind historical series with ARIMA forward forecast
#| echo: false
#| results: hide
#| message: false
#| code-fold: true
#| cache: true
```

```{r}
#| label: g1
#| code-summary: Hero graph — historical series + 24-month ARIMA projection
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "Full historical series (Apr 2005 – Sep 2025, steelblue) with ARIMA 24-month forward projection (blue). Shaded regions: 80% (darker) and 95% (lighter) prediction intervals. Dashed line marks latest observed data point."
print(g1_forecast_report)
```

## Key Forecast Numbers

The table below anchors the visual in concrete numbers. Projections are from the **ARIMA (Tier 2)** model at +6, +12, +18, and +24 months from the focal date, with 95% prediction intervals.

```{r}
#| label: t1
#| code-summary: Point forecasts and 95% prediction intervals at key horizons
#| echo: false
#| code-fold: true
#| cache: true
t1_data %>%
  knitr::kable(
    caption = paste0(
      "ARIMA (Tier 2) point forecasts at selected horizons. ",
      "Change vs Now = change from September 2025 (", fmt_num(current_caseload), " cases)."
    ),
    align = c("r", "l", "l", "r", "r", "r", "r", "r")
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  )
```

---

# Model Comparison

Two forecasting models are estimated in this phase. They represent opposite ends of the model-complexity spectrum: the **Seasonal Naive** defines the floor of acceptable performance (any useful model must beat it), while **ARIMA** captures temporal autocorrelation and trend dynamics.

## Faceted Forecasts

Each panel below isolates one model at full size, with its own uncertainty envelope. The shared y-axis enables a direct comparison of forecast levels and interval widths.

```{r}
#| label: g2-data-prep
#| code-summary: Combine both tiers for comparison plots
#| echo: false
#| results: hide
#| message: false
#| code-fold: true
#| cache: true
```

```{r}
#| label: g2
#| code-summary: Faceted forecast — one panel per model tier
#| code-fold: true
#| eval: true
#| cache: true
#| fig-height: 8
#| fig-cap: "Oct 2025 – Sep 2027 projections for each model tier. Tier 1 (coral): repeats last year's monthly pattern with constant level. Tier 2 (blue): ARIMA (3,1,1)(1,0,0)[12] captures trend momentum. ARIMA prediction intervals are systematically narrower because log-transformation stabilises forecast variance."
print(g2_faceted_forecast)
```

## Model Divergence

Overlaying the two models on a single panel reveals *how fast* the gap opens as the horizon extends.

```{r}
#| label: g21
#| code-summary: Overlaid forecast — divergence between ARIMA and Naive
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "Both models overlaid. ARIMA (blue) trends upward, driven by the estimated AR(3) autoregression on the log-transformed series. Seasonal Naive (coral) anchors on the most-recent 12-month cycle, producing a flat seasonal pattern. The divergence grows from ~500 cases at +6 months to several thousand at +24 months."
print(g21_overlaid_forecast)
```

## Performance Metrics

The table below summarises every available performance metric from the 24-month held-out backtest window (October 2023 – September 2025). **Lower is better for all metrics.**

```{r}
#| label: t2
#| code-summary: RMSE, MAE, and MAPE for all model tiers
#| echo: false
#| code-fold: true
#| cache: true
t2_data %>%
  knitr::kable(
    caption = paste0(
      "Backtest performance (Oct 2023 – Sep 2025, 24 months). ",
      "RMSE/MAE in absolute caseload; MAPE in percent. ",
      "All metrics computed on back-transformed (original) scale. ",
      "ARIMA order shown for Tier 2; AICc not applicable (N/A) for rule-based Tier 1."
    )
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  ) %>%
  kableExtra::row_spec(
    which(t2_data$MAPE == min(t2_data$MAPE)),
    bold = TRUE, background = "#e8f4f8"
  )
```

**Reading the table**: ARIMA reduces MAPE by approximately 3 percentage points relative to the Seasonal Naive — from ~16% to ~13%. In absolute terms, the ARIMA model errs by roughly 700 fewer cases per month on average (MAE difference). This is meaningful for budget planning: 700 cases × average benefit cost = non-trivial fiscal exposure.

---

# Backtest Evidence

This section shows *how* the models performed month-by-month during the held-out period, not just *how well* (as summarised by the metrics above). Month-level diagnostics reveal systematic biases that summary statistics obscure.

> **Important note on "backtest" interpretation**: The fitted values displayed here are *one-step in-sample predictions* from models estimated on the full 246-month series. True multi-step hold-out metrics (RMSE/MAE/MAPE in the performance table) were computed separately in Lane 4 (`4-train-IS.R`) by training on the 222-month training slice and forecasting the final 24 months forward. The two approaches are complementary; this section uses the in-sample diagnostics for visual residual analysis.

## Actual vs Model Fitted

```{r}
#| label: g3-data-prep
#| code-summary: Merge backtest comparison data for diagnostic plots
#| echo: false
#| results: hide
#| message: false
#| code-fold: true
#| cache: true
```

```{r}
#| label: g3
#| code-summary: Actual vs fitted — 24-month backtest window
#| code-fold: true
#| eval: true
#| cache: true
#| fig-height: 8
#| fig-cap: "Black solid line = actual caseload. Coloured dashed line = model-fitted values. Both models track the general level reasonably well; ARIMA tracks month-to-month movements more closely because it learns autocorrelation structure."
print(g3_actual_vs_fitted)
```

## Residual Pattern

A well-specified model's residuals should be random — no systematic structure, no persistent sign. Coloured bars below zero indicate months where the model over-predicted (it said caseload would be higher than it was), and bars above zero indicate under-prediction.

```{r}
#| label: g31
#| code-summary: Residuals over the backtest window
#| code-fold: true
#| eval: true
#| cache: true
#| fig-height: 8
#| fig-cap: "Residual = actual – fitted. Blue bars (positive): model under-predicted caseload. Coral bars (negative): model over-predicted. Seasonal Naive shows a persistent negative bias in later months — it over-predicts because it anchors on the COVID-era rebound year, which was atypically high. ARIMA residuals are more centred around zero."
print(g31_residuals)
```

## Percentage Error by Month

Absolute percentage error reveals *proportional* accuracy — months where the model missed by the most relative to actual caseload, regardless of scale.

```{r}
#| label: g32
#| code-summary: Absolute percentage error vs MAPE reference line
#| code-fold: true
#| eval: true
#| cache: true
#| fig-height: 8
#| fig-cap: "Each point = absolute monthly percentage error. Dashed line = MAPE (mean across 24 months). Points above the dashed line are worse-than-average months. Both models struggle most around fiscal year transitions (Apr–May) and at the beginning of the backtest window (Oct 2023), when the COVID tail effects were still dissipating."
print(g32_pct_error)
```

---

# What the Evidence Suggests: Policy Implications

The forecasting results carry several implications for caseload planning and policy:

**1. Upward pressure is real, but modest.** The ARIMA model's central forecast shows a gradual upward trend over the 24-month horizon. This reflects the autoregressive momentum in the series — the recent uptick since 2023 has not reversed and the model projects it to continue at a subdued pace. It does *not* predict a sudden spike.

**2. Prediction intervals are wide — plan for ranges, not point estimates.** The 95% interval at +24 months spans thousands of cases. Budget analysts should prepare high, central, and low scenarios corresponding to the upper bound of the 80% interval, the point forecast, and the lower bound respectively. Relying on a single point forecast for multi-year budget commitments underestimates fiscal risk.

**3. The Seasonal Naive over-predicted during the backtest.** Persistent negative residuals indicate the naive model anchored on the COVID-era caseload rebound year (Oct 2022 – Sep 2023), which was atypically elevated. This systematic bias renders Tier 1 unsuitable as a standalone policy forecast — it is retained only as an accuracy benchmark.

**4. ARIMA does not explain *why* caseload moves.** The current ARIMA specification is univariate — it extrapolates temporal patterns without incorporating economic drivers (unemployment rate, oil price, wage growth, policy changes). This limits its usefulness for scenario analysis. *What happens to caseload if unemployment rises 2 percentage points?* The current models cannot answer this.

**5. The path to narrower uncertainty and richer scenario capability runs through Tiers 3 and 4.** Adding client-type composition as a static regressor (Tier 3) will reduce model error on the demographic dimension. Adding a time-varying economic predictor (Tier 4 — unemployment rate or oil price are natural candidates given Alberta's economic structure) will enable counter-factual scenario analysis essential for strategic planning.

**6. Model retraining cadence matters.** Both models are estimated on data through September 2025. As new monthly data arrives, the models should be re-estimated (rerun Lanes 3–5) to maintain forecast freshness. The pipeline is designed for this: changing `focal_date` in `config.yml` propagates automatically through all downstream lanes.

---

# Model Specifications

Technical specifications for analysts who need to validate or extend the modelling work.

```{r}
#| label: t3
#| code-summary: Model registry — specifications and information criteria
#| echo: false
#| code-fold: true
#| cache: true
t3_data %>%
  select(-`Forge Hash`) %>%
  knitr::kable(
    caption = paste0(
      "Model specification registry (from model_registry.csv). ",
      "AIC/AICc/BIC: information criteria — lower values indicate better fit penalised for complexity; ",
      "not defined for rule-based models (Tier 1). ",
      "Forge Hash (first 8 chars): links model to its exact Mint data slice."
    )
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  )
```

**Reading the ARIMA order `(3,1,1)(1,0,0)[12]`:**

| Component | Symbol | Value | Meaning |
|-----------|--------|-------|---------|
| Non-seasonal AR | p | 3 | Caseload depends on the previous 3 months (after differencing) |
| Non-seasonal differencing | d | 1 | One regular difference needed to achieve stationarity |
| Non-seasonal MA | q | 1 | One lagged forecast error in the model |
| Seasonal AR | P | 1 | One seasonal autoregressive term at lag 12 |
| Seasonal differencing | D | 0 | No seasonal differencing needed |
| Seasonal MA | Q | 0 | No seasonal moving average |
| Seasonality period | [s] | 12 | Monthly data with annual seasonality |

All ARIMA orders were selected automatically by `forecast::auto.arima(stepwise = FALSE, approximation = FALSE)` on the log-transformed series, consistent with EDA decisions [EDA-001] through [EDA-003] codified in Lane 3 (Mint).

---

# Data Lineage and Provenance

Full traceability from raw data to forecast output. Any authorised user can reproduce these results by running the pipeline from `flow.R` with the same `config.yml`.

## Provenance Summary (Manifests)

```{r}
#| label: inspect-data-0
#| code-summary: Full provenance dashboard — manifests, lineage check, registry snapshot
#| code-fold: true
#| eval: true
#| cache: true
```

## Lane 3 (Mint) — Data Contract

The `forge_manifest.yml` records the exact data decisions codified in Lane 3. These decisions derive from EDA-2 diagnostics and are the authoritative record of what went into model training.

```{r}
#| label: t4
#| code-summary: forge_manifest.yml contents (Lane 3 data contract)
#| echo: false
#| code-fold: true
#| cache: true
t4_data %>%
  knitr::kable(
    col.names = c("Field", "Value"),
    caption   = "forge_manifest.yml — The data contract between Lane 3 (Mint) and Lane 4 (Train). forge_hash links every model and forecast back to this exact data slice."
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width        = FALSE
  ) %>%
  kableExtra::column_spec(1, bold = TRUE, width = "12em") %>%
  kableExtra::column_spec(2, width = "28em")
```

## Lane 5 (Forecast) — Execution Record

```{r}
#| label: t5
#| code-summary: forecast_manifest.yml contents (Lane 5 provenance)
#| echo: false
#| code-fold: true
#| cache: true
t5_data %>%
  knitr::kable(
    col.names = c("Field", "Value"),
    caption   = "forecast_manifest.yml — Execution record for Lane 5 (Forecast). forecast_hash is a content fingerprint of all forecast output files. forge_hash_consumed links back to the Mint artifacts consumed by the Train models."
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width        = FALSE
  ) %>%
  kableExtra::column_spec(1, bold = TRUE, width = "12em") %>%
  kableExtra::column_spec(2, width = "28em")
```

### Pipeline Execution Order

The full reproducible pipeline runs via `Rscript flow.R` from the project root:

| Lane | Script | Pattern | Role |
|------|--------|---------|------|
| 1 | `manipulation/1-ferry.R` | Ferry | Import raw CSV from Open Alberta |
| 2 | `manipulation/2-ellis.R` | Ellis | Transform to 11 analysis-ready parquet tables |
| — | `analysis/eda-2/eda-2.qmd` | EDA | Diagnose trends, stationarity, transform decisions |
| 3 | `manipulation/3-mint-IS.R` | Mint | Train/test split, log transform, xreg matrices |
| 4 | `manipulation/4-train-IS.R` | Train | Fit Tier 1 (snaive) + Tier 2 (ARIMA) on full series |
| 5 | `manipulation/5-forecast-IS.R` | Forecast | Generate 24-month projections + backtest diagnostics |
| **6** | **`analysis/report-1/report-1.qmd`** | **Report** | **This document** |

---

# Environment

> Technical appendix. Non-technical readers are welcome to skip; return here to understand the origins of custom functions, packages, and data objects.

**Packages used** (see `load-packages` chunk sourced via `read_chunk()`): `magrittr`, `ggplot2`, `forcats`, `stringr`, `lubridate`, `dplyr`, `tidyr`, `scales`, `arrow`, `yaml`, `knitr`, `kableExtra`, `forecast`, `fs`.

**Custom functions** sourced from:

- [`scripts/common-functions.R`](../../scripts/common-functions.R) — `neat()`, `quick_save()`, `safe_numeric_convert()`
- [`scripts/operational-functions.R`](../../scripts/operational-functions.R) — `compute_fiscal_year()`, `month_diff()`, etc.

**Local helpers** defined in `declare-functions`: `format_fiscal_year_label()`, `fmt_pct()`, `fmt_num()`, `make_tier_label()`.

**Config** (`config.yml`): `focal_date = 2025-09-01`, `backtest_months = 24`, `forecast_horizon = 24`, `use_log_transform = true`, `seasonal_period = 12`, `random_seed = 42`.

**Colour constants**: `steelblue` (historical series), `#2166ac` (ARIMA / Tier 2), `#d6604d` (Seasonal Naive / Tier 1), `gray50` (split-date reference line).

All graphing conventions follow [`analysis/eda-1/eda-style-guide.md`](../eda-1/eda-style-guide.md): dual-file pattern (`.R` for development with `ggsave()`, `.qmd` for publication with `print()`), default dimensions 10 × 6 inches at 300 DPI, `theme_minimal()` globally applied.

---

# Session Info

```{r}
#| label: session-info
#| code-summary: R session information for reproducibility
#| echo: false
#| code-fold: true
cat("Report completed at:", as.character(Sys.time()), "\n")
cat("Total runtime:", round(difftime(Sys.time(), report_render_start_time, units = "mins"), 2), "minutes\n")
sessionInfo()
```
