---
title: "Alberta Income Support Caseload: Exploratory Data Analysis"
subtitle: "EDA 2 - Time Series Analysis of Total Caseload and Client Types"
author: "Strategic Data Analytics Team"
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: show
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
    keep-md: false
editor: visual
editor_options: 
  chunk_output_type: console
---

# Mission

This document provides an **exploratory analysis** of Alberta Income Support caseload data, introducing readers to:

-   **Total Caseload Trends**: 20-year historical perspective (April 2005 - September 2025)
-   **Client Type Composition**: How caseload breaks down by employment readiness categories
-   **Temporal Patterns**: Seasonal effects, year-over-year changes, and policy impact periods
-   **Data Preparation for Forecasting**: Understanding historical patterns to inform predictive models

**Key Analytical Questions:**
 - What are the long-term trends in Alberta Income Support caseload? 
 - How did major economic events (2008 crisis, oil price collapse, COVID-19) affect caseload? 
 - How has client composition shifted over time (Expected to Work vs Barrier-Free Employment)? 
 - What seasonal patterns exist in the data?
 - What growth rates characterize different historical periods?

**Data Sources:**
- **Raw Data**: Alberta Open Data - Income Support Monthly Caseload Statistics
- **Analysis-Ready Data**: Processed through Ellis pattern ([manipulation/2-ellis.R](../../manipulation/2-ellis.R))
- **Documentation**: See [CACHE-manifest.md](../../data-public/metadata/CACHE-manifest.md) for complete data dictionary

Following the **template workflow**, we develop analytical insights iteratively in the companion `.R` script before publishing them here for broader consumption.

# **Definition of Terms**

```{r}
#| code-summary: Reconciles working directory between execution modes
#| echo: false
#| results: hide
#| message: false 
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir='../../') # keep this chunk isolated
```

```{r}
#| label: set_options
#| code-summary: report-wide formatting options
#| echo: false
#| results: hide
#| message: false 
cat("Working directory: ", getwd()) # turn on to test the location
report_render_start_time <- Sys.time()
# set options shared by all chunks
opts_chunk$set(
  results      = 'show',
  message      = FALSE,
  warning      = FALSE,
  comment      = NA,
  tidy         = FALSE,
  # dpi        = 400, # dots per inch,
  out.width  = "960px", # pixels, this affects only the markdown, not the underlying png file.  The height will be scaled appropriately.
  fig.width    = 10, # inches - larger for publication trends
  fig.height   = 6, # inches - good for time series
  fig.path     = 'figure-png-iso/' # where figures are stored
  
)
echo_chunks    <- FALSE #Toggle for debugging.
message_chunks <- FALSE #Toggle for debugging.
options(width=100) # number of characters to display in the output (dflt = 80)
ggplot2::theme_set(ggplot2::theme_minimal()) # clean theme for publications
read_chunk("analysis/eda-2/eda-2.R") #This allows knitr to call chunks from R script
```

# Environment

> Reviews the components of the working environment of the report. Non-technical readers are welcomed to skip. Come back if you need to understand the origins of custom functions, scripts, or data objects.

```{r}
#| label: load-packages
#| code-summary: Packages used in the current report
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-sources
#| code-summary: Collection of custom functions used in current repository
#| echo: true
#| results: hide
#| message: false
#| code-fold: true

```

```{r}
#| label: declare-globals
#| code-summary: Values used throughout the report.
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-functions
#| code-summary: Custom functions defined for use in this report.
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

# Data

This report analyzes two primary datasets from the analysis-ready CACHE outputs:

-   **ds0_total → ds1_total**: Complete time series from April 2005 to September 2025 (246 months) - the longest historical view available
-   **ds0_client_type → ds1_client_type**: Client type breakdown from April 2012 to September 2025 (162 months) - introduced with Alberta Works reform

**Data Genealogy:** 
- Raw data → [Ferry Lane 1](../../manipulation/1-ferry.R) → CACHE database 
- CACHE staging → [Ellis Lane 2](../../manipulation/2-ellis.R) → Analysis-ready parquet files
- **Loaded as**: ds0_total, ds0_client_type
- **Transformed to**: ds1_total, ds1_client_type (date formatting applied)
- **Data Quality**: All Ellis quality checks passed, suppressed values handled, Nov 2020 duplicate corrected

**Client Type Categories:**
- **ETW: Working** - Employed but income insufficient for self-sufficiency
- **ETW: Available for Work** - Job-ready, actively seeking employment  
- **ETW: Unavailable for Work** - Temporarily unable (caregiving, illness, training)
- **Barrier-Free Employment** - Significant barriers (disabilities, complex circumstances)

```{r }
#| label: load-data
#| code-summary: Load analysis-ready data (ds0_total, ds0_client_type from parquet)
#| warning: false
#| code-fold: true
#| cache: true
```

```{r}
#| label: tweak-data-1
#| code-summary: Transform to ds1_total and ds1_client_type with proper date formatting
#| code-fold: true
#| eval: true
#| cache: true
```

```{r}
#| label: inspect-data-0
#| code-summary: Loaded data overview (ds0_total, ds0_client_type)
#| code-fold: false
#| eval: true
#| cache: true
```

```{r}
#| label: inspect-data-1
#| code-summary: Total caseload structure (ds1_total after transformation)
#| code-fold: false
#| eval: true
#| cache: true
```

```{r}
#| label: inspect-data-2
#| code-summary: Total caseload summary statistics (ds1_total)
#| code-fold: false
#| eval: true
#| cache: true
```

```{r}
#| label: inspect-data-3
#| code-summary: Client type data structure (ds1_client_type after transformation)
#| code-fold: false
#| eval: true
#| cache: true
```

```{r}
#| label: inspect-data-4
#| code-summary: Client type categories and coverage (ds1_client_type)
#| code-fold: false
#| eval: true
#| cache: true
```

# Analysis

## Long-Term Caseload Trends

### Total Caseload Time Series (2005-2025)

The complete 20-year historical perspective shows major inflection points driven by economic shocks and policy changes:

```{r}
#| label: g1
#| code-summary: Total caseload time series from April 2005 to September 2025
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "Alberta Income Support total caseload over 246 months - full historical context"
```

**Key Observations:**
- **2008-09 Financial Crisis**: Sharp increase in caseload as unemployment rose
- **2012 Alberta Works Reform**: Policy changes introduced Expected to Work framework
- **2014-16 Oil Price Collapse**: Sustained increase reflecting Alberta's resource economy sensitivity
- **COVID-19 Pandemic (2020-21)**: Dramatic spike followed by rapid decline as federal programs absorbed demand
- **Post-2022 Trends**: Stabilization at pre-pandemic baseline levels

---

### Historical Period Comparison

Aggregate statistics by major policy and economic periods:

```{r}
#| label: g2-data-prep
#| code-summary: Prepare period-based summary statistics
#| code-fold: true
#| eval: true
#| cache: true
```

```{r}
#| label: g2
#| code-summary: Average caseload by historical period
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "How average monthly caseload varied across seven distinct historical periods"
```

**Implications for Forecasting:**
- Each period has distinct baseline levels and volatility patterns
- COVID-19 represents unprecedented shock but relatively brief duration
- Oil price sensitivity suggests economic indicators as valuable predictors
- Policy changes (2012 reform) created structural shifts in caseload composition

---

## Client Type Composition Analysis

### Stacked Area Chart: Client Type Evolution

How caseload composition has shifted since the introduction of the Expected to Work framework in 2012:

```{r}
#| label: g3
#| code-summary: Client type composition as stacked area chart
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "Evolution of client type mix from April 2012 to September 2025 - stacked view"
```

**Key Insights:**
- **Barrier-Free Employment** (red) forms a relatively stable base layer
- **ETW: Available for Work** (light blue) shows highest volatility - most responsive to economic conditions
- **ETW: Working** (dark blue) represents income supplementation - reflects working poverty dynamics
- **ETW: Unavailable for Work** (medium blue) smallest category, relatively stable

---

### Faceted Time Series: Individual Category Trajectories

Detailed view of each client type's independent trajectory:

```{r}
#| label: g4
#| code-summary: Client type trends shown in separate facets with independent y-axes
#| code-fold: true
#| eval: true
#| cache: true
#| fig-height: 8
#| fig-cap: "Four separate time series allowing comparison of growth patterns and volatility"
```

**Forecasting Strategy Implications:**
- Each category exhibits distinct patterns - **model separately rather than aggregating**
- **ETW: Available for Work** most sensitive to economic indicators (unemployment rate, job postings)
- **Barrier-Free Employment** more stable - likely driven by demographic factors and disability policy
- **ETW: Working** tied to wage growth, minimum wage policy, and working poverty rates

---

## Temporal Patterns and Seasonality

### Year-over-Year Comparison (Recent Years)

Monthly patterns across 2020-2025 showing seasonal and pandemic effects:

```{r}
#| label: g5-data-prep
#| code-summary: Prepare year-over-year comparison data
#| code-fold: true
#| eval: true
#| cache: true
```

```{r}
#| label: g5
#| code-summary: Monthly caseload by year (2020-2025)
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "Overlay of monthly patterns for recent years - seasonality and pandemic shock visible"
```

**Seasonality Observations:**
- **Spring increases** (March-May) suggest fiscal year-end effects or seasonal employment transitions
- **Summer dips** (June-August) may reflect seasonal employment opportunities
- **Fall increases** (September-November) could relate to school year transitions
- **2020-21 pandemic years** broke typical seasonal patterns

---

### Growth Rate Analysis

Year-over-year percentage changes reveal volatility and structural shifts:

```{r}
#| label: g6-data-prep
#| code-summary: Calculate month-over-month and year-over-year growth rates
#| code-fold: true
#| eval: true
#| cache: true
```

```{r}
#| label: g6
#| code-summary: Year-over-year percentage change time series
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "YoY growth rate showing periods of increase (red) and decrease (green)"
```

**Growth Rate Insights:**
- **High volatility periods**: 2008-09 crisis, 2014-16 oil collapse, 2020-21 pandemic
- **Peak growth**: +40-50% YoY during crisis peaks
- **Sustained declines**: Post-peak recovery periods often show 20-30% YoY drops
- **Baseline volatility**: ±5-10% YoY in "stable" periods suggests ongoing sensitivity to economic conditions

---

# Conclusions and Next Steps

## Key Findings

1. **Long-term trends** show Alberta Income Support caseload is highly sensitive to economic shocks, particularly oil price movements and unemployment rate changes

2. **Client type composition** has remained relatively stable since 2012 reform, with ETW: Available for Work showing highest volatility

3. **Seasonal patterns** exist but are modest compared to economic cycle effects - seasonality should be included in models but is not dominant

4. **COVID-19 pandemic** created unprecedented but relatively brief shock - federal programs (CERB, CRB) absorbed demand that might otherwise have flowed to provincial Income Support

5. **Growth rate volatility** is substantial - forecasting models must account for regime changes and structural breaks

## Recommendations for Forecasting Work

1. **Model client types separately** - each category has distinct drivers and volatility profiles

2. **Include economic predictors** - unemployment rate, oil price, wage growth, job postings

3. **Account for policy changes** - 2012 reform, COVID-19 federal programs represent structural breaks

4. **Use ensemble methods** - given high uncertainty and regime changes, combine multiple model approaches

5. **Wide prediction intervals** - conservative uncertainty estimates given historical volatility

6. **Regular model retraining** - structural changes require frequent model updates with recent data

## Next Analysis Steps

- **EDA 3**: Analyze family composition and regional breakdown tables
- **EDA 4**: External predictor variables (economic indicators, demographic trends)
- **Model Development**: Begin with ARIMA baseline, progress to multivariate time series models
- **Azure ML Integration**: Set up AutoML experiments for ensemble forecasting

---

# Session Info

```{r}
#| label: session-info
#| code-summary: R session information for reproducibility
#| echo: false
#| code-fold: true
cat("Report completed at:", as.character(Sys.time()), "\n")
cat("Total runtime:", round(difftime(Sys.time(), report_render_start_time, units = "mins"), 2), "minutes\n")
sessionInfo()
```
