---
title: "Data Primer — {NAME}"
subtitle: "Canonical data documentation: Ellis outputs, schemas, variable reference"
author: "Report Composer (Composing Orchestra)"
date: last-modified
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: show
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
editor: visual
editor_options:
  chunk_output_type: console
---

<!-- Composing Orchestra data primer template.
     Used for analysis/data-primer-1/ — the CENTRALIZED data reference for the project.
     Composed once via @report-composer, verified by humans, linked by all EDAs and Reports.
     Customize {PLACEHOLDERS} and populate sections from Ellis metadata + CACHE manifest. -->

# Purpose

This is the **canonical data primer** for the project. It provides the foundational documentation of all data assets produced by the Ellis lane. Every EDA and presentation Report in `analysis/` links here for data context.

This primer answers: *What data do we have? What does it look like? What do the variables mean?*

For domain terminology, see `ai/project/glossary.md`.

```{r}
#| code-summary: Reconciles working directory
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir='../../')
```

```{r}
#| label: setup
#| code-summary: Environment setup
#| echo: false
#| results: hide
#| message: false
library(magrittr)
library(dplyr)
library(tidyr)
library(arrow)
library(knitr)
library(scales)

opts_chunk$set(
  results   = 'show',
  message   = FALSE,
  warning   = FALSE,
  comment   = NA,
  tidy      = FALSE,
  out.width = "960px"
)
options(width = 100)
```

# Pipeline Architecture

The data assets used in this project follow a three-stage provenance chain:

```
Source data (data-private/raw/)
  → Ellis lane (manipulation/2-ellis.R)
    → R-friendly parquet files (data-private/derived/)
```

**Ellis** imports raw source data, applies R-friendly type conversions (factors for categoricals, dates, integers for IDs), validates against the CACHE manifest, and outputs optimized parquet files.

# Data Assets Inventory

```{r}
#| label: inventory
#| code-summary: List available parquet files from Ellis output
#| code-fold: true

# List Ellis outputs
ellis_path <- "./data-private/derived/"
if (dir.exists(ellis_path)) {
  parquet_files <- list.files(ellis_path, pattern = "\\.parquet$", full.names = TRUE, recursive = TRUE)
  if (length(parquet_files) > 0) {
    inventory <- data.frame(
      File = basename(parquet_files),
      Path = parquet_files,
      stringsAsFactors = FALSE
    )
    # Read schema info for each file
    inventory$Rows <- NA_integer_
    inventory$Columns <- NA_integer_
    for (i in seq_along(parquet_files)) {
      tryCatch({
        ds_temp <- arrow::read_parquet(parquet_files[i])
        inventory$Rows[i] <- nrow(ds_temp)
        inventory$Columns[i] <- ncol(ds_temp)
      }, error = function(e) {
        message("Could not read: ", parquet_files[i])
      })
    }
    kable(inventory[, c("File", "Rows", "Columns")], caption = "Ellis Output: Analysis-Ready Parquet Files")
  } else {
    cat("No parquet files found in:", ellis_path, "\n")
    cat("Run the Ellis lane first.\n")
  }
} else {
  cat("Ellis output directory not found at:", ellis_path, "\n")
  cat("Run the Ellis lane first.\n")
}
```

# Table Schemas

<!-- Customize these sections based on the actual Ellis outputs for this project.
     Remove or add sections as needed to match the available parquet files. -->

## Primary Dataset

```{r}
#| label: schema-primary
#| code-summary: Primary dataset structure
#| code-fold: true

# Update path to match your project's actual Ellis output
path_primary <- paste0(ellis_path, "{PRIMARY_TABLE}.parquet")
if (file.exists(path_primary)) {
  ds_primary <- arrow::read_parquet(path_primary)
  cat("Grain: {GRAIN_DESCRIPTION}\n")
  cat("Rows:", scales::comma(nrow(ds_primary)), "| Columns:", ncol(ds_primary), "\n\n")
  glimpse(ds_primary)
} else {
  cat("File not found:", path_primary, "\n")
}
```

# Grain Demonstration

```{r}
#| label: grain-check
#| code-summary: Verify the grain of each table
#| code-fold: true

# Update grain check to match your project's key columns
# if (exists("ds_primary") && is.data.frame(ds_primary)) {
#   cat("--- Primary Dataset ---\n")
#   n_total <- nrow(ds_primary)
#   n_unique <- ds_primary %>% distinct({KEY_COL1}, {KEY_COL2}) %>% nrow()
#   cat("Total rows:", scales::comma(n_total), "\n")
#   cat("Unique {KEY_COL1} x {KEY_COL2}:", scales::comma(n_unique), "\n")
#   cat("Grain holds (1:1):", n_total == n_unique, "\n\n")
# }
```

# Single Record View

```{r}
#| label: single-record
#| code-summary: What the data looks like for one representative record
#| code-fold: true

# Update to show a representative record from your primary dataset
# if (exists("ds_primary") && is.data.frame(ds_primary)) {
#   cat("=== Example record ===\n\n")
#   ds_primary %>% slice_sample(n = 1) %>% glimpse()
# }
```

# Variable Reference

Key variables, sourced from `data-public/metadata/CACHE-manifest.md`:

<!-- Populate this section during the adaptive interview with the @report-composer.
     Update column names, types, and descriptions to match your actual dataset. -->

## Identifiers

| Variable | Type | Description |
|----------|------|-------------|
| `{id_col}` | {type} | {description} |

## Key Dimensions

| Variable | Type | Description |
|----------|------|-------------|
| `{dim_col}` | {type} | {description} |

## Key Measures

| Variable | Type | Description |
|----------|------|-------------|
| `{measure_col}` | {type} | {description} |

# CACHE Manifest Summary

For complete documentation of data coverage, quality metrics, and variable availability, see `data-public/metadata/CACHE-manifest.md`.
